\chapter{Preface}

This is a collection of my reading notes on Hilbert and Bernays's
\emph{Grundlagen der Mathematiks}, vol.\ I (1934). I have decided to
order my notes by paragraph (counted within each chapter), using the
German 1970 edition.

I am prefixing ``\pilcrow'' to the paragraph numbers, unindented and
in bold, prefixed by the chapter. That is, ``\pilcrow$\langle c\rangle.\langle p\rangle$''
where $c$ is the chapter number and $p$ is the paragraph number.

Furthermore, when one paragraph develops multiple points of
discussion, I add a ``subparagraph'' number. That is to say, I write
``\pilcrow$\langle c\rangle.\langle p\rangle.\langle s\rangle$''
where $s$ is the subparagraph number. When there are subparagraphs, it
is numbered within each paragraph.

The chapter, section, and subsection numbers follows Hilbert and
Bernays's conventions. They look like
``\S$\langle c\rangle$\texttt{(}$\langle s\rangle$\texttt{)}\texttt{(}$\langle t\rangle$\texttt{)}''
where $s$ is the section number written as a lowercase Latin letter,
$t$ is the subsection number. I don't like this typography, but it is
\emph{their} convention (\emph{not mine!\/}).

\paragraph*{English translation}
I am relying heavily on Claus-Peter Wirth's translation of the first
few chapters of the first volume, because the German is too eloquent
for my meagre comprehension.

Further, Ian Mueller~\cite{mueller2006grundlagen} wrote an unpublished
draft translating the first two chapters back around 2006. I rely on
Mueller to give some sense where there is a completely different
reading of the same passages.

